---
title: "Assignment 2 - Language Development in ASD - Making predictions"
author: "Riccardo Fusaroli"
date: "August 9, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome to the second exciting part of the Language Development in ASD exercise

In this exercise we will delve more in depth with different practices of model comparison and model selection, by first evaluating your models from last time against some new data. Does the model generalize well?
Then we will learn to do better by cross-validating models and systematically compare them.

The questions to be answered (in a separate document) are:
1- Discuss the differences in performance of your model in training and testing data
2- Which individual differences should be included in a model that maximizes your ability to explain/predict new data?
3- Predict a new kid's performance (let's call him Bernie) and discuss it against expected performance of the two groups

## Learning objectives

- Critically appraise the predictive framework (contrasted to the explanatory framework)
- Learn the basics of machine learning workflows: training/testing, cross-validation, feature selections

## Let's go

N.B. There are several datasets for this exercise, so pay attention to which one you are using!

1. The (training) dataset from last time (the awesome one you produced :-) ).
2. The (test) datasets on which you can test the models from last time:
* Demographic and clinical data: https://www.dropbox.com/s/5pc05mh5jwvdfjk/demo_test.csv?dl=0
* Utterance Length data: https://www.dropbox.com/s/eegu8fea2entdqv/LU_test.csv?dl=0
* Word data: https://www.dropbox.com/s/cf4p84mzn2p1bev/token_test.csv?dl=0

### Exercise 1) Testing model performance

How did your models from last time perform? In this exercise you have to compare the results on the training data and on the test data. Report both of them. Compare them. Discuss why they are different.

- recreate the models you chose last time (just write the model code again and apply it to your training data (from the first assignment))
- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())
- create the test dataset (apply the code from assignment 1 to clean up the 3 test datasets)
- test the performance of the models on the test data (Tips: google the functions "predict()")
- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())




```{r, include = FALSE}
pacman::p_load(readr, dplyr, stringr, lmerTest, Metrics, caret, MuMIn,  lme4)

```


```{r, include = FALSE}
## Loading test data
Demo <- read_csv("demo_test.csv")
LU <- read_csv("LU_test.csv")
Word <- read_csv("token_test.csv")

## Clean up function, (using this later)

CleanUpData <- function(Demo,LU,Word){
  
  Speech <- merge(LU, Word) %>% 
    rename(
      Child.ID = SUBJ, 
      Visit=VISIT) %>%
    mutate(
      Visit = as.numeric(str_extract(Visit, "\\d")),
      Child.ID = gsub("\\.","", Child.ID)
      ) %>%
    dplyr::select(
      Child.ID, Visit, MOT_MLU, CHI_MLU, types_MOT, types_CHI, tokens_MOT, tokens_CHI
    )
  
  Demo <- Demo %>%
    dplyr::select(
      Child.ID, Visit, Ethnicity, Diagnosis, Gender, Age, ADOS, MullenRaw, ExpressiveLangRaw, Socialization
    ) %>%
    mutate(
      Child.ID = gsub("\\.","", Child.ID)
    )
    
  Data=merge(Demo,Speech,all=T)
  
  Data1= Data %>% 
     subset(Visit=="1") %>% 
     dplyr::select(Child.ID, ADOS, ExpressiveLangRaw, MullenRaw, Socialization) %>%
     rename(Ados1 = ADOS, 
            verbalIQ1 = ExpressiveLangRaw, 
            nonVerbalIQ1 = MullenRaw,
            Socialization1 = Socialization) 
  
  Data=merge(Data, Data1, all=T) %>%
    mutate(
      Child.ID = as.numeric(as.factor(as.character(Child.ID))),
      Visit = as.numeric(as.character(Visit)),
      Gender = recode(Gender, 
         "1" = "M",
         "2" = "F"),
      Diagnosis = recode(Diagnosis,
         "A"  = "ASD",
         "B"  = "TD")
    )

  return(Data)
}
```


```{r setting up data and columns}
# Loading training Data
df <- read_csv("data_clean.csv")


# Removing NAs from certain columns
completeFun <- function(data, desiredCols) {
  completeVec <- complete.cases(data[, desiredCols])
  return(data[completeVec, ])
}

df <- completeFun(df, "CHI_MLU")
df <- completeFun(df, "Age")

df <- completeFun(df, "verbalIQ1")
df <- completeFun(df, "nonVerbalIQ1")
df <- completeFun(df, "Socialization1")
```



```{r setting up data and columns}
# creating ADOS2 and other variable columns without NA, but copied from visit 1

# subsetting the 6 comlumns 
#subset <- dplyr::select(df, Visit, Child.ID, Visit, ADOS, MullenRaw, ExpressiveLangRaw, Socialization)
# only including visit 1 in the subset
#subset <- filter(subset, Visit == 1)
# deleting the visit column
#subset <- dplyr::select(subset, -Visit)
# rename clinical measures variables
#subset <- rename(subset, ADOS2 = ADOS)
#subset <- rename(subset, MullenRaw2 = MullenRaw)
#subset <- rename(subset, ExpressiveLangRaw2 = ExpressiveLangRaw)
#subset <- rename(subset, Socialization2 = Socialization)
# merging the subset with df
#df <- merge(df, subset, by = "Child.ID")



# MullenRaw = Non-verbal IQ
# ExpressiveLangRaw = Verbal IQ
```



```{r load models and view r.squared}
#- recreate the models you chose last time (just write the code again and apply it to Train Data)
# The models: best quad for h1, last one with age instead of VISIT

model <- lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + verbalIQ1 + (1 + Diagnosis + MOT_MLU | Child.ID), df)


```


```{r evaluating performance of model using RMSE() function}
#- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the function rmse())

# Making a list of the model predictions:
predictions_a2 <- predict(model, df, allow.new.level = TRUE)
# Comparing model predictions to actual observations:
RMSE(df$CHI_MLU, predictions_a2, na.rm = TRUE)
# The average residual (difference between observed and predicted value) is 0.29 from the observed traning data to the model predictions.

# For context, the mean CHI_MLU in the training data is 1.99 (the RMSE is in the same original measurement unit)
mean(df$CHI_MLU, na.rm = T)




############ Should we perhaps in this one use fitted() instead of predict(), because we are calculating the performance of the model on the training data (i.e. the data that the model is based on)? Also hint: It is not until later that se asks us to google predict(), so maybe we shouldnt use it here.

# Removing the rows where the CHI_MLU is NA
completeFun <- function(data, desiredCols) {
  completeVec <- complete.cases(data[, desiredCols])
  return(data[completeVec, ])
}

#df <- completeFun(df, "CHI_MLU")
#df <- completeFun(df, "Age")

# Making a list of the model fit:
fit <- fitted(model)
# Comparing model predictions to actual observations:
RMSE(df$CHI_MLU, fit, na.rm = TRUE)
# The average residual (difference between observed and predicted value) is 0.29 from the observed traning data to the model predictions.

# For context, the mean CHI_MLU in the training data is 1.99 (the RMSE is in the same original measurement unit)
mean(df$CHI_MLU, na.rm = T)
```


```{r}
#- create the test dataset (apply the code from assignment 1 or my function to clean up the 3 test datasets)
# loading test datasets
Demo <- read_csv("demo_test.csv")
LU <- read_csv("LU_test.csv")
Word <- read_csv("token_test.csv")
#trying to use the CleanUpData() function on them
CleanTest_df <- CleanUpData(Demo, LU, Word)


# renaming some variable names to fit the clean traning data
CleanTest_df <- rename(CleanTest_df, ADOS1 = Ados1)

```


```{r}
#- test the performance of the models on the test data (Tips: google the functions "predict()")
predictions_againstTrain <- predict(model, CleanTest_df, allow.new.levels = TRUE)
RMSE(CleanTest_df$CHI_MLU, predictions_againstTrain, na.rm = TRUE)

# Here, the average residual AKA the root mean squared error is 0.41 from the observed TEST data to the model predictions
# For context, the mean observed CHI_MLU in the test data is 2.1 (the RMSE is in the same original measurement unit)
mean(CleanTest_df$CHI_MLU, na.rm = T)
```





```{r}
#- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())

predictInterval(
  model,
  CleanTest_df,
  which = c("full", "fixed", "random", "all"),
  level = 0.8,
  n.sims = 1000,
  stat = c("median", "mean"),
  type = c("linear.prediction", "probability"),
  include.resid.var = TRUE,
  returnSims = FALSE,
  seed = NULL,
  .parallel = FALSE,
  .paropts = NULL,
  fix.intercept.variance = FALSE,
  ignore.fixed.terms = NULL
)
```


'''

[HERE GOES YOUR ANSWER]

### Exercise 2) Model Selection via Cross-validation (N.B: ChildMLU!)

One way to reduce bad surprises when testing a model on new data is to train the model via cross-validation. 

In this exercise you have to use cross-validation to calculate the predictive error of your models and use this predictive error to select the best possible model.

- Use cross-validation to compare your model from last week with the basic model (Child MLU as a function of Time and Diagnosis, and don't forget the random effects!)
- (Tips): google the function "createFolds";  loop through each fold, train both models on the other folds and test them on the fold)

```{r}
# Basic model:
#basic <- lmer(CHI_MLU ~ Visit*Diagnosis + (1 + Diagnosis + MOT_MLU | Child.ID), df)


# For the basic model:
k <-  10
folds <- createFolds(unique(df$Child.ID), k = k, list = TRUE, returnTrain = FALSE)
trainRMSE_basic <- rep(NA, k)
testRMSE_basic <- rep(NA, k)
i <- 1

for (fold in folds){
  train = subset(df, !(Child.ID %in% fold))
  test = subset(df, Child.ID %in% fold)
  model = lmer(CHI_MLU ~ Visit*Diagnosis + (1 + Diagnosis + MOT_MLU | Child.ID), train)
  test$prediction <- predict(model, test, allow.new.levels = TRUE)
  train$prediction <- fitted(model)
  trainRMSE_basic[i] <- RMSE(train$CHI_MLU, fitted(model))
  testRMSE_basic[i] <- RMSE(test$CHI_MLU, test$prediction)
  i <- i + 1
}

trainRMSE_basic
testRMSE_basic

# For the other model:
k <-  10
folds <- createFolds(unique(df$Child.ID), k = k, list = TRUE, returnTrain = FALSE)
trainRMSE_m <- rep(NA, k)
testRMSE_m <- rep(NA, k)
i <- 1

for (fold in folds){
  train = subset(df, !(Child.ID %in% fold))
  test = subset(df, Child.ID %in% fold)
  model = lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + verbalIQ1 + (1 + Diagnosis + MOT_MLU | Child.ID), train)
  test$prediction <- predict(model, test, allow.new.levels = TRUE)
  train$prediction <- fitted(model)
  trainRMSE_m[i] <- RMSE(train$CHI_MLU, fitted(model))
  testRMSE_m[i] <- RMSE(test$CHI_MLU, test$prediction)
  i <- i + 1
}

trainRMSE_m
testRMSE_m

crossValidation <- data.frame(trainRMSE_basic, testRMSE_basic, trainRMSE_m, testRMSE_m)

mean_train <- c(mean(crossValidation$trainRMSE_basic), mean(crossValidation$trainRMSE_m))

mean_test <- c(mean(crossValidation$testRMSE_basic), mean(crossValidation$testRMSE_m))

models <- c("basic", "m")

crossValidationMean <- data.frame(models, mean_train, mean_test) # m wins

```


- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

- Bonus Question 1: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?
- Bonus Question 2: compare the cross-validated predictive error against the actual predictive error on the test data


```{r}
#- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.


# Models to test in the first round (with varying random slopes):
# m <- lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + verbalIQ1 + (1 + Diagnosis + MOT_MLU | Child.ID), train)
# m1 <- lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + verbalIQ1 + (1 | Child.ID), train)
# m2 <- lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + verbalIQ1 + (1 +  MOT_MLU | Child.ID), train)
# m3 <- lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + verbalIQ1 + (1 + Diagnosis | Child.ID), train)


######### m1 ##########

k <-  10
folds <- createFolds(unique(df$Child.ID), k = k, list = TRUE, returnTrain = FALSE)
trainRMSE_m1 <- rep(NA, k)
testRMSE_m1 <- rep(NA, k)
i <- 1

for (fold in folds){
  train = subset(df, !(Child.ID %in% fold))
  test = subset(df, Child.ID %in% fold)
  model = m1 <- lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + verbalIQ1 + (1 | Child.ID), train)
  test$prediction <- predict(model, test, allow.new.levels = TRUE)
  train$prediction <- fitted(model)
  trainRMSE_m1[i] <- RMSE(train$CHI_MLU, fitted(model))
  testRMSE_m1[i] <- RMSE(test$CHI_MLU, test$prediction)
  i <- i + 1
}

trainRMSE_m1
testRMSE_m1

######### m2 ##########

k <-  10
folds <- createFolds(unique(df$Child.ID), k = k, list = TRUE, returnTrain = FALSE)
trainRMSE_m2 <- rep(NA, k)
testRMSE_m2 <- rep(NA, k)
i <- 1

for (fold in folds){
  train = subset(df, !(Child.ID %in% fold))
  test = subset(df, Child.ID %in% fold)
  model = lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + verbalIQ1 + (1 +  MOT_MLU | Child.ID), train)
  test$prediction <- predict(model, test, allow.new.levels = TRUE)
  train$prediction <- fitted(model)
  trainRMSE_m2[i] <- RMSE(train$CHI_MLU, fitted(model))
  testRMSE_m2[i] <- RMSE(test$CHI_MLU, test$prediction)
  i <- i + 1
}

trainRMSE_m2
testRMSE_m2

######### m3 ##########

k <-  10
folds <- createFolds(unique(df$Child.ID), k = k, list = TRUE, returnTrain = FALSE)
trainRMSE_m3 <- rep(NA, k)
testRMSE_m3 <- rep(NA, k)
i <- 1

for (fold in folds){
  train = subset(df, !(Child.ID %in% fold))
  test = subset(df, Child.ID %in% fold)
  model = lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + verbalIQ1 + (1 + Diagnosis | Child.ID), train)
  test$prediction <- predict(model, test, allow.new.levels = TRUE)
  train$prediction <- fitted(model)
  trainRMSE_m3[i] <- RMSE(train$CHI_MLU, fitted(model))
  testRMSE_m3[i] <- RMSE(test$CHI_MLU, test$prediction)
  i <- i + 1
}

trainRMSE_m3
testRMSE_m3



crossValidation1 <- data.frame(trainRMSE_m, testRMSE_m, trainRMSE_m1, testRMSE_m1, trainRMSE_m2, testRMSE_m2, trainRMSE_m3, testRMSE_m3)


mean_train <- c(mean(crossValidation1$trainRMSE_m), mean(crossValidation1$trainRMSE_m1), mean(crossValidation1$trainRMSE_m2), mean(crossValidation1$trainRMSE_m3))

mean_test <- c(mean(crossValidation1$testRMSE_m), mean(crossValidation1$testRMSE_m1), mean(crossValidation1$testRMSE_m2), mean(crossValidation1$testRMSE_m3))

models <- c("m", "m1", "m2", "m3")

crossValidation1Mean <- data.frame(models, mean_train, mean_test) # m still wins

##### m4

k <-  10
folds <- createFolds(unique(df$Child.ID), k = k, list = TRUE, returnTrain = FALSE)
trainRMSE_m4 <- rep(NA, k)
testRMSE_m4 <- rep(NA, k)
i <- 1

for (fold in folds){
  train = subset(df, !(Child.ID %in% fold))
  test = subset(df, Child.ID %in% fold)
  model = lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + (1 + Diagnosis | Child.ID), train)
  test$prediction <- predict(model, test, allow.new.levels = TRUE)
  train$prediction <- fitted(model)
  trainRMSE_m4[i] <- RMSE(train$CHI_MLU, fitted(model))
  testRMSE_m4[i] <- RMSE(test$CHI_MLU, test$prediction)
  i <- i + 1
}


##### m5

k <-  10
folds <- createFolds(unique(df$Child.ID), k = k, list = TRUE, returnTrain = FALSE)
trainRMSE_m5 <- rep(NA, k)
testRMSE_m5 <- rep(NA, k)
i <- 1

for (fold in folds){
  train = subset(df, !(Child.ID %in% fold))
  test = subset(df, Child.ID %in% fold)
  model = lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + nonVerbalIQ1 + (1 + Diagnosis | Child.ID), train)
  test$prediction <- predict(model, test, allow.new.levels = TRUE)
  train$prediction <- fitted(model)
  trainRMSE_m5[i] <- RMSE(train$CHI_MLU, fitted(model))
  testRMSE_m5[i] <- RMSE(test$CHI_MLU, test$prediction)
  i <- i + 1
}

##### m6 

k <-  10
folds <- createFolds(unique(df$Child.ID), k = k, list = TRUE, returnTrain = FALSE)
trainRMSE_m6 <- rep(NA, k)
testRMSE_m6 <- rep(NA, k)
i <- 1

for (fold in folds){
  train = subset(df, !(Child.ID %in% fold))
  test = subset(df, Child.ID %in% fold)
  model = lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + Socialization1 + (1 + Diagnosis | Child.ID), train)
  test$prediction <- predict(model, test, allow.new.levels = TRUE)
  train$prediction <- fitted(model)
  trainRMSE_m6[i] <- RMSE(train$CHI_MLU, fitted(model))
  testRMSE_m6[i] <- RMSE(test$CHI_MLU, test$prediction)
  i <- i + 1
}

crossValidation2 <- data.frame(trainRMSE_m3, testRMSE_m3, trainRMSE_m4, testRMSE_m4, trainRMSE_m5, testRMSE_m5, trainRMSE_m6, testRMSE_m6)


mean_train <- c(mean(crossValidation2$trainRMSE_m3), mean(crossValidation2$trainRMSE_m4), mean(crossValidation2$trainRMSE_m5), mean(crossValidation2$trainRMSE_m6))

mean_test <- c(mean(crossValidation2$trainRMSE_m3), mean(crossValidation2$testRMSE_m4), mean(crossValidation2$testRMSE_m5), mean(crossValidation2$testRMSE_m6))

models <- c("m3", "m4", "m5", "m6")

crossValidation2Mean <- data.frame(models, mean_train, mean_test) # m still wins





# Bonus Question 1: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?
# Bonus Question 2: compare the cross-validated predictive error against the actual predictive error on the test data
```

[HERE GOES YOUR ANSWER]

### Exercise 3) Assessing the single child

Let's get to business. This new kiddo - Bernie - has entered your clinic. This child has to be assessed according to his group's average and his expected development.

Bernie is one of the six kids in the test dataset, so make sure to extract that child alone for the following analysis.

You want to evaluate:

- how does the child fare in ChildMLU compared to the average TD child at each visit? Define the distance in terms of absolute difference between this Child and the average TD.

```{r}
# Making a df with only data for Bernie
bernie <- filter(CleanTest_df, Child.ID == "2")

# Making a df with only data for TD children
td <- filter(df, Diagnosis == "TD")

# Finding the mean CHI_MLU of the TD children for each visit
MLU_mean_perVisit <- data.frame(aggregate(td[, 13], list(td$Visit), mean)) #"Group.1" = visit and "x" = CHI_MLU mean
MLU_mean_perVisit <- rename(MLU_mean_perVisit, Visit = Group.1)

# Adding a column to the Bernie data frame with the 
bernie$diff <- bernie$CHI_MLU - MLU_mean_perVisit$CHI_MLU

# Difference across visits:
mean(bernie$CHI_MLU) - mean(MLU_mean_perVisit$CHI_MLU)
```

- how does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations? (tip: use the predict() function on Bernie's data only and compare the prediction with the actual performance of the child)

```{r}
# Our model (m3):
m3 <- lmer(CHI_MLU ~ Visit + I(Visit^2) + Diagnosis + Visit*Diagnosis + verbalIQ1 + (1 + Diagnosis | Child.ID), df)

# Making a list of the model predictions:
predict_bernie <- predict(m1, bernie, allow.new.level = TRUE)

# Comparing model predictions to actual observations:
RMSE(bernie$CHI_MLU, predict_bernie, na.rm = TRUE)
# The average residual (difference between observed and predicted value) is 0.46 from the observed traning data to the model predictions.

# For context, the mean CHI_MLU for TD children in the training data is 2.31 (the RMSE is in the same original measurement unit)
mean(td$CHI_MLU, na.rm = T)


```


[HERE GOES YOUR ANSWER]

### OPTIONAL: Exercise 4) Model Selection via Information Criteria
Another way to reduce the bad surprises when testing a model on new data is to pay close attention to the relative information criteria between the models you are comparing. Let's learn how to do that!

Re-create a selection of possible models explaining ChildMLU (the ones you tested for exercise 2, but now trained on the full dataset and not cross-validated).

Then try to find the best possible predictive model of ChildMLU, that is, the one that produces the lowest information criterion.

- Bonus question for the optional exercise: are information criteria correlated with cross-validated RMSE? That is, if you take AIC for Model 1, Model 2 and Model 3, do they co-vary with their cross-validated RMSE?

### OPTIONAL: Exercise 5): Using Lasso for model selection
Welcome to the last secret exercise. If you have already solved the previous exercises, and still there's not enough for you, you can expand your expertise by learning about penalizations. Check out this tutorial: http://machinelearningmastery.com/penalized-regression-in-r/ and make sure to google what penalization is, with a focus on L1 and L2-norms. Then try them on your data!

